---
title: "BatChef: A common interface for batch effect correction methods in single-cell RNA-seq data"
author: 
- name: Elena Zuin
  affiliation: Department of Biology, University of Padova, Italy
- name: Chiara Romualdi
  affiliation: Department of Biology, University of Padova, Italy
- name: Davide Risso
  affiliation: Department of Statistical Sciences, University of Padova, Italy
- name: Gabriele Sales
  affiliation: Department of Biology, University of Padova, Italy
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: BatChef
vignette: >
  %\VignetteIndexEntry{batch_correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE,
    out.width = "100%"
)
```

# Introduction

Combining single-cell RNA-sequencing (scRNA-seq) datasets from different sources 
introduces technical variability known as batch effects, caused by factors like 
operators, reagent batches, platforms, and timing. These batch effects distort 
analyses and obscure true biological signals, making accurate data interpretation difficult.

In the literature there are various batch correction methods, based on different
mathematical approaches, using R and Python languages.

`BatChef` is an R package that provides a common interface for batch effect 
correction methods in single-cell RNA-seq data. It allows users to compute 
various correction methods through a generic function by specifying the desired
approach. The package contains multiple methods implemented in both R and Python,
each based on different mathematical approach.

Also, it provides metrics to evaluate the performance of the correction methods, 
such as the Wasserstein distance, Local Inverse Simpson's Index (LISI), Average 
Silhouette Width (ASW), and Adjusted Rand Index (ARI).

# Installation

```{r,eval=FALSE}
# install.packages("devtools")
devtools::install_github("zuinelena3/BatChef")
```

# Setting up

```{r, include=TRUE, results="hide", message=FALSE, warning=FALSE}
#Load the packages
library(BatChef)
library(Seurat)
library(reticulate)
library(zellkonverter)
```

To demonstrate, we simulate a dataset with 2,000 genes, 2,000 cells, 
2 batches and 2 cell types, using the `simulated_data` function. This function 
applies the same parameters as the Splatter package’s ones. 
Additionally, it normalizes the data, identifies highly variable genes via 
the `scrapper` package, and performs principal component analysis using 
the `scater` package.

```{r}
sce <- simulated_data(nGenes = 2000, batchCells = c(800, 1200), 
                      group.prob = c(0.4, 0.6), n_hvgs = 1000, 
                      ncomp = 10, seed = 3)
sce
```

# Batch effects correction

The `batchCorrect` function allows to specify the desired correction method and 
performs the correction with three different input data: `SingleCellExperiment`,
`Seurat` and `AnnData`. 
So, we convert our data.

```{r, message=FALSE, warning=FALSE}
so <- as.Seurat(sce)
```

```{r, eval=FALSE}
# Error: C stack usage  7969796 is too close to the limit
# andata <- zellkonverter::SCE2AnnData(sce)
```


The desired correction method can have these values:

-   `LimmaParams`

-   `CombatParams`

-   `Seuratv3Params`

-   `Seuratv5Params`

-   `FastMNNParams`

-   `HarmonyParams`

-   `ScanoramaParams`

-   `ScVIParams`

-   `ScMerge2Params`

-   `BBKNNParams`

-   `LigerParams`

## Examples 

To illustrate, we correct the data using `Scanorama` and `Harmony`.

### Harmony

```{r, message=FALSE, warning=FALSE}
sce <- batchCorrect(input = sce, batch = "Batch", params = HarmonyParams())
sce
```

Also, the `Seurat` object can be used as an input data.

```{r, message=FALSE, warning=FALSE}
so <- batchCorrect(input = so, batch = "Batch", params = HarmonyParams())
so
```

### Scanorama

```{r, message=FALSE, warning=FALSE}
sce <- batchCorrect(input = sce, batch = "Batch", params = ScanoramaParams(assay_type = "logcounts",
                                                                           return_dimred = TRUE))
sce
so <- batchCorrect(input = so, batch = "Batch", params = ScanoramaParams(assay_type = "data", 
                                                                         return_dimred = TRUE))
so
```

# Performance evaluation

Last, the `metrics`()` function computes the evaluation metrics, such as 
Normalized Mutual Information (NMI), Adjusted Rand Index (ARI), 
Local Inverse Simpson's Index (LISI), Average Silhouette Width (ASW), and 
Wasserstein distance.

The Wasserstein distance is computed on a resampled data due to the different 
number of cells in each batch. The parameter `rep` indicates how many times 
the calculation should be repeated (in this case 10).

```{r, message=FALSE, warning=FALSE}
red <- SingleCellExperiment::reducedDimNames(sce)
metrics <- lapply(red, function(x) metrics(input = sce, batch = "Batch",
                                           group = "Group", reduction = x, 
                                           rep = 5))
metrics <- do.call(rbind, metrics)
metrics
```

```{r}
sessionInfo()
```

